Nitrogen's form determines its fate in the North Pacific Subtropical Gyre during PARAGON I

### 2025-05-12 manuscript/manuscript.Rmd
Pivoting results section to be xamend first

Nitrogen availability controls productivity in much of the surface ocean. The form of this nitrogen controls who it is available to and how it is used. Here, we explored how four different forms of nitrogen (ammonia, nitrate, arginine, and guanosine monophosphate) were taken up and transformed in the North Pacific Subtropical Gyre using stable isotope labeling metabolomics. We repeated these experiments both at the surface and at the base of the euphotic zone as well as in the morning and evening. We found that ammonia is converted into glutamate in less than 20 minutes and results in labeling throughout the metabolome, though osmolytes like betaines had slow turnover rates. This process was slower at depth but independent of time of day. Nitrate amendments resulted in the labeling of only a small subset of molecules such as mycosporine-like amino acids, whose incorporation patterns suggest that this form of nitrogen was available to a subset of the community, likely eukaryotes. The two organic nitrogen amendments labeled the metabolome more slowly than the inorganic forms of nitrogen and were primarily converted into glutamate rather than used as intact substrates. They also showed the same usage at depth as they did at the surface, hinting at the major role that heterotrophic bacteria likely play in their uptake and use. These measurements reveal the forms and fluxes of organic nitrogen in the natural environment to provide constraints on biogeochemical models and better characterize the transformations between organic and inorganic nitrogen.

>>>>

Nitrogen availability controls productivity in much of the surface ocean and the form of this nitrogen controls who it is available to and how it is used. Here, we explored how four different forms of nitrogen (ammonia, nitrate, arginine, and guanosine monophosphate) were taken up and transformed in the North Pacific Subtropical Gyre using stable isotope labeling metabolomics. We repeated these experiments both at the surface and at the base of the euphotic zone as well as in the morning and evening. We found that ammonia is converted into glutamate in less than 20 minutes and results in labeling throughout the metabolome, though osmolytes like betaines had slow turnover rates. This process was slower at depth but largely independent of time of day. Nitrate amendments resulted in the labeling of only a small subset of molecules such as mycosporine-like amino acids, whose incorporation patterns suggest that this form of nitrogen was available to a subset of the community, likely eukaryotes. The two organic nitrogen amendments labeled the metabolome more slowly than the inorganic forms of nitrogen and showed similar usage at depth and at the surface, hinting at the major role that heterotrophic bacteria likely play in their uptake and use. A large fraction of the organic nitrogen added was remineralized into ammonia and used as glutamate rather than as intact substrates, though labeled arginine was incorporated into proteins directly. These measurements reveal the forms and fluxes of organic nitrogen in the natural environment to provide constraints on biogeochemical models and better characterize the transformations between organic and inorganic nitrogen.
